Acceleration of plague outbreaks in the second pandemic

### 2020-07-12 supp.Rnw
supp now compiles with PNAS SI format

\documentclass[12pt,table]{article}
%% knitr requires documentclass to be the first line of the file
%% Created: 19 Mar 2017
%% Changed: 22 Apr 2017
%%          30 Apr 2017
\input{nlipreamble.tex}

@@

%%\definecolor[named]{verylightgray}{rgb}{0.95,0.95,0.95}
\definecolor[named]{verylightgray}{rgb}{0.93,0.93,0.93}

\newcommand{\mysection}[1]{\clearpage\section*{#1}\addcontentsline{toc}{section}{#1}}
\newcommand{\mysectionsamepage}[1]{\section*{#1}\addcontentsline{toc}{section}{#1}}
%% http://tex.stackexchange.com/questions/316510/hyperlink-within-caption-to-a-figure
%% if used in a caption, need short content \caption[...]{ ... \Methodslink ... }
\newcommand{\Methodslink}{\href[page=9]{ms.pdf}{\emph{Methods}}}
\newcommand\ttbackslash{{\tt\char`\\}}
\newcommand{\macro}[1]{{\tt\ttbackslash#1}}

@@

%%\title{Faster growth of later plagues in the second millenium\\\emph{Electronic Supplementary Material}}
\title{Acceleration of plague outbreaks in the second
  pandemic\\\emph{Electronic \supp}}

\author{David J.\,D.\ Earn$^{a,b,c}$, 
Junling Ma$^{f}$, 
Hendrik N.\ Poinar$^{b,c,d,e}$,\\
Jonathan Dushoff$^{a,b,c}$,
and Benjamin M.\ Bolker$^{a,b,c}$\\
\href{mailto:earn@math.mcmaster.ca}{\tt earn@math.mcmaster.ca}\\
\small $^a$Department of Mathematics \& Statistics, $^b$Department of Biology, \\
\small $^c$Michael G.\ deGroote Institute for Infectious Disease Research,\\
\small $^d$McMaster Ancient DNA Centre, Department of Anthropology,\\
\small and $^e$Department of Biochemistry,\\
\small McMaster University, 1280 Main Street West, Hamilton, ON, L8S 4K1, Canada;\\
\small $^f$Department of Mathematics \& Statistics, University of Victoria, Victoria, BC, Canada}

@@

\linenumbers

\null\vspace{-1.5cm}

\vbox{
\maketitle

\tableofcontents
}

\section{Historical human population estimates for London}

%
\begin{table}[h!]
\caption{Population estimates for London previously published by Finlay \cite{Finl81,FinlShea86}\cite[p.\,108]{Kryl11} and Creighton \cite[p.\,660]{Crei65}.  These data are plotted in \cref{F:london_pop}.}
\label{tab:pop}
\begin{center}
%% \input{analysis/tables/supppop.tex} % generated by analysis/tables/supppop.R
%% DE: I'm building the (somewhat different) table myself.
\renewcommand{\arraystretch}{1.0}
\begin{tabular}{c|r|l}
%%\hline
{\bfseries Year} &
{\bfseries Population} &
{\bfseries Source}\\
\hline
  1340 &  45,000 & Finlay \\ 
  1500 &  50,000 & Finlay \\ 
  1550 & 110,000 & Finlay \\ 
  1600 & 180,000 & Finlay \\ 
  1603 & 250,000 & Creighton \\
  1625 & 320,000 & Creighton \\
  1650 & 340,000 & Finlay \\ 
  1665 & 460,000 & Creighton \\
  1700 & 445,000 & Finlay \\ 
\end{tabular}
\end{center}
\end{table}
%

%
\begin{figure}[h!]
\begin{center}
\includegraphics[width=\textwidth]{analysis/plots/london_pop.pdf}
\end{center}
\caption{Historical population of London, England, as estimated by Finlay \cite{Finl81,FinlShea86} (\cref{tab:pop}).  Earlier estimates of Creighton \cite[p.\,660]{Crei65} are shown for the specific plague years 1603, 1625 and 1665.}
\label{F:london_pop}
\end{figure}
%

\FloatBarrier
\section{Disease-related data sources}

@@

%% requires tabularx package:
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\begin{table}[h!]
\caption[]{Disease-related data used in this paper.}
\label{tab:datasources}
\begin{center}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{c|c|L{3cm}|c|L{3.5cm}}
%%\hline
{\bfseries Epoch} &
{\bfseries Year range} &
{\bfseries Data type} &
{\bfseries Frequency} &
{\bfseries Source} %%&
%%{\bfseries Description}
 \\
\hline
Early & 1340--1380 & Last Wills and Testaments & Daily & Court of Husting \\\hline
Late & 1540--1680 & Last Wills and Testaments & Daily & Prerogative Court of Canterbury (PCC) \\
 & 1540--1680 & Mortality from all causes & Weekly & Parish records\\
 & 1563--1666 & Mortality from plague & Weekly & London Bills of Mortality (LBoM)
\end{tabular}
\end{center}
\end{table}

\subsection{Last wills and testaments}

@@

\subsection{Will dates vs.\ probate dates}

@@

%
\begin{figure}[h!]
\begin{center}
  \includegraphics[width=\textwidth]{analysis/plots/probate_dates.pdf}
\end{center}
\caption{Counts of wills written vs.\ wills probated during each of
  the four 14th century plague epidemics in London, based on wills
  probated in the Court of Husting \cite{Shar1889}. Smooth lines represent loess fits
with span=0.5 (shaded regions are 95\% confidence intervals).}
\label{F:willsVSprobates}
\end{figure}
%

\subsection{London Bills of Mortality (LBoM)}

@@

\subsection{Parish registers}

@@

\subsection{Major vs minor plague epidemics}

@@

%%\newcommand{\mywidth}{0.8\textwidth}
\newcommand{\mywidth}{\textwidth}
%
\begin{figure}[h!]
\begin{center}
\includegraphics[width=\mywidth]{analysis/plots/plotallscaled.crop.pdf}
\end{center}
\caption{Plague deaths per 1000 persons per week, from 1563 to 1666.
Epidemics that exceeded 5 on this scale were classified as \emph{major}.
Analysis of the major epidemics is shown in \cref{tab:params,F:timeseries,F:combdata}.  Corresponding analyses of all of the minor epidemics are presented
in \cref{tab:minorparams,F:seasonalgrowth}.
}
\label{F:all1600s.scaled}
\end{figure}
%

\begin{figure}[h!]
\begin{center}
%%\scalebox{0.6}{
\includegraphics[width=\mywidth]{analysis/plots/plotall.crop.pdf}
%%}
\end{center}
\caption{All plague deaths reported in London from 1563 to 1666.}
\label{F:all1600s}
\end{figure}
%

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{analysis/plots/plotwills14c.pdf}
\end{center}
\caption[]{Weekly numbers of wills written during plague epidemics in London, England, in the 14th century.
Vertical grey lines indicate the ``outbreak window'' specified in \cref{tab:windows}.}
\label{F:plotwills14c}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=\textwidth]{analysis/plots/plot3sourcesmajors.pdf}
\end{center}
\caption[]{Weekly deaths and numbers of wills written during major plague epidemics in London, England, since 1540.  For each epidemic, vertical grey lines indicate the beginning and end of the period during which deaths from plague were listed in the bills of mortality.  The left (right) scale refers to deaths (wills).}
\label{F:plot3sourcesmajors}
\end{figure}

%%\clearpage
\section{Generation interval for bubonic plague}\label{sec:gentime}

For pneumonic plague, sufficiently detailed data exist for a number of
modern outbreaks to allow an estimate of the latent and infectious
periods (and hence of the generation interval distribution)
\cite{GaniLeac04}, which we can use to estimate $\rzero$ for a given
value of $r$ \cite{WallLips07}. Much less data is available for bubonic plague; in addition, the more complex host-vector life cycle of bubonic plague complicates the estimation of the generation interval. However, we can say that the elapsed time between the onset of infectiousness of a rat and the time when a rat in the next infection cycle becomes infectious is:
%%
\begin{equation}
  \begin{split}
    &\quad \textrm{rat} \to \textrm{flea infection time} \\
    \mbox{} + &\quad \textrm{max}\bigl( \textrm{flea incubation period},
    \{\textrm{time to rat death} + \textrm{flea searching time}\} \bigr) \\
    \mbox{} + &\quad \textrm{flea} \to \textrm{rat infection time} \\
    \mbox{} + &\quad \textrm{rat incubation period}
  \end{split}
\end{equation}
%%
The second line above takes account of the fact that a flea leaves
its rat host when, and only when, the rat dies \cite{KeelGill00}.  From
\cite{LaPerriere+2009}, we can gather that
\begin{itemize}
\item fleas bite rats $\approx$ 4 times/day; the rat $\to$ flea infection probability is $\approx$ 0.2 and the flea $\to$ rat infection probability is $\approx$ 0.28, suggesting that both the rat $\to$ flea and flea $\to$ rat infection times are $\approx$ 1 day; 
\item the rat infectious period (time until infected rats die) is $\approx$ 4 days (\cite{LaPerriere+2009} cite \cite{Audi03} for this value; \cite{KeelGill00b} give a value of $\approx$ 18 days, which seems unrealistically long --- another source, \cite{Toll+10}, gives values ranging from $\approx$ 4--9 days)
\item the flea incubation period is long (but very variable), ranging from 9--26 days (\cite{LaPerriere+2009}, citing \cite{Poll52})
\item the flea searching time is not explicitly defined by \cite{LaPerriere+2009}, but we guess it is relatively short ($<$ 1--2 days)
\item \cite{LaPerriere+2009} give a value of $\approx$ 1--3 days for the rat incubation period
\end{itemize}
%%
Since the flea incubation period is typically longer than the combination of rat death time and flea searching time (9--26 days vs.\ (4--9 + 1--2) days), we can approximate the generation interval as $\approx$ (1 day) + (9--26 days) + (1 day) + (1--3 days); we use a value of $\gentime=18$ days in the main text.

\section{Relationship between rat density and $\R_0$ (in rats)}

@@

%
\begin{figure}[h!]
\begin{center}
 \includegraphics[width=\textwidth]{analysis/plots/KvsR.pdf}
\end{center}
\caption{Relative change in rat carrying capacity $K_{\rm R}$ as a
  function of relative change in basic reproduction number $\R_0$.
  Solid curves are based on \cref{E:R0ratio}.  The dotted red line is
  based on \cref{E:aKsmall}.  The \Sexpr{round(R0ratio,2)}-fold increase in $\R_0$ that
  requires explanation is indicated in yellow.}
\label{F:KvsR}
\end{figure}
%

\input{analysis/fits/modeldefs.tex}
\section{Effects of rat ecology on growth rates}

@@

As a vector-borne disease, the rate of bubonic plague spread is primarily affected by the flea-rat ratio.
In the traditional Ross-MacDonald model for vector-borne epidemics, $\rzero$ is proportional to the square of vector-host ratio because vectors must independently bite hosts twice (once to become infected and a second time to infect a susceptible host). In the rat-flea model, $\rzero$ is instead proportional to $V/H$ because \emph{all} infected fleas on a host disperse and bite other hosts when their initial host dies.
From Keeling and Gilligan's plague model \cite{KeelGill00b}, the
expected change in $r$ for a change in flea-rat ratio from $K_{\rm
  F0}$ to $K_{\rm F}$ is $(K_{\rm F}/K_{\rm F0})^S$, where $S$ is the
sensitivity ($\approx 1.5$). Thus in order to a see a \foldval
increase in $r$ we would need a change of $4^{(1/S)} \approx \Sexpr{round(4^(1/1.5),1)}$, not in the rat density, but in the number of fleas per rat, which seems unlikely.

As noted in the main text, rat density has an additional, indirect effect on growth rate $r$ and reproduction number $\rzero$: increasing rat density will increase the probability that fleas leaving dying rats will find new, susceptible rat hosts. To crudely estimate the magnitude of this effect, if we again consider Keeling and Gilligan's model \cite{KeelGill00b} then the maximum possible effect of rat density on $\rzero$ by this mechanism would be proportional (if fleas have very low success in finding new rat hosts). That is, in order for $\rzerob$ to increase from $\approx \rzeroearlybest$ in the 14th century to $\approx \rzerolatebest$ in the late epoch, rat density would have to increase by at least $\ratpct\%$. If fleas were already fairly successful at finding hosts in the 14th century, then changes in rat density would be expected to have only a small effect on $\rzero$ in rats.

\section{Analysis of delays between wills and other sources}

@@

\begin{table}[h!]
\caption{Observed cross-correlation and delay between London Bills of
  Mortality (LBoM) records and other sources. All lags measured in
  weeks.  Rows corresponding to wills data are highlighted in grey.
}
\begin{center}
  \input{analysis/tables/wills_delay_tab.tex}
\end{center}
\label{tab:delays}
\end{table}

\section{Details of numerical optimization}

Robustly fitting phenomenological epidemic curves to small, noisy data
sets proved to be surprisingly challenging\footnote{See ``monsters in
  the basement'',
  \url{https://redpenblackpen.tumblr.com/post/145304820562/monsters-in-the-basement}.}. In
the course of developing the full model we experimented with a range
of optimizers --- specifically the BFGS and Nelder-Mead options for
R's \code{optim} function as well as R's \code{nlminb} function; we
also tried an approach that iterated back and forth between
Nelder-Mead and BFGS until convergence was achieved or the fit was
sufficiently stable. We chose \code{nlminb} because it gave the best
results (highest log likelihood) for point estimates.
In general, we used the standard link functions proposed by Ma \emph{et al.} \cite{Ma+14}, i.e., fitting most of the parameters (growth rate $r$, background mortality/wills rate $b$, final size $K$, Richards shape parameter $s$) on the log scale with a scaled-tangent link ($\eta = \tan(\pi/2 \cdot (x/2-1))$) for the initial condition $x_0$. However, we found that the scaled-tangent link declined too slowly in the tails, leading to numerical instability; we replaced it with a more standard logit link ($\eta = \log(x/(1-x))$).

@@

\FloatBarrier
%%\clearpage
\section{Supplementary tables cited in the main text}

\begin{table}[h!]
    \cprotect\caption{Summary statistics for the model estimating differences across epochs. All parameters are in units of log(growth rate)/year.
      The model includes fixed effects of epoch (early [14th c.]
      vs. late [16th - 17th c.]) and source (wills, parish, London
      Bills of Mortality) and a random effect of outbreak
      year. Variability for each observation is assumed to be
      proportional to the uncertainty in its $\log(r)$ estimate (see
      main text, \emph{Growth rate estimates}). 
      In Wilkinson-Rogers notation, the model formula is: \verb!log.r ~ epoch + source + (1|outbreak.year), disp= ~ 1+offset(log(sdvals^2))!. Parameter estimates are given on the $\log(r)$ scale;
      Wald confidence intervals are given in parentheses.
}  \begin{center}
    \input{analysis/tables/epochsum.tex}
  \end{center}
\label{tab:summarystats}
\end{table}

\begin{table}
  \caption[Minor outbreak parameter estimates]{Parameter estimates for minor plague epidemics. Plague epidemics with $\hat r<1/\mathrm{yr}$ or $\hat r>100/\mathrm{yr}$, representing unreliable fits, are excluded; confidence limits that are $<0.1$ or $>200$ are replaced with {\tt NA}. See \Methodslink.}
  \footnotesize
  \rowcolors{2}{white}{verylightgray}
  \begin{center}
    \input{analysis/tables/minorparams.tex}
  \end{center}
  \label{tab:minorparams}
\end{table}

\begin{table}
  \caption[Fitting windows]{Outbreak windows and fitting windows used in fitting the major plague epidemics.  See \Methodslink.}
  \rowcolors{2}{white}{verylightgray}
  \input{analysis/tables/suppwindows.tex}
  \label{tab:windows}
\end{table}

\begin{table}
  \caption[Fitting windows (minors)]{The start and end dates for the fitting windows used in fitting the minor plague epidemics.  See \Methodslink.}
  \scriptsize
  \rowcolors{2}{white}{verylightgray}
  \begin{center}
    \input{analysis/tables/suppwindowsminor.tex}
  \end{center}
  \label{tab:windowsminor}
\end{table}

\FloatBarrier
\section{Supplementary figure cited in the main text}

%
\begin{figure}[h!]
\begin{center}
\scalebox{0.9}{
  \includegraphics{analysis/plots/seasonalgrowth.pdf}
}
\end{center}
\caption{Growth rate estimates for all the late epoch (1540--1680) London plague epidemics.  The growth rates shown for the major epidemics are the same as those in the right panel of \cref{F:combdata}. 
}
\label{F:seasonalgrowth}
\end{figure}
%

%%\clearpage
\bibliographystyle{pnas-new}
\bibliography{plague}
\addcontentsline{toc}{section}{References}

>>>>

\documentclass[9pt,twocolumn,twoside,lineno,table]{pnas-new}
% Use the lineno option to display guide line numbers if required.
%% knitr requires documentclass to be the first line of the file
%% Created: 19 Mar 2017
%% Changed: 22 Apr 2017
%%          30 Apr 2017

@@

\usepackage{xspace}
\usepackage{cprotect} % allow verbatim in caption
\usepackage{hhline} % Better horizontal lines in tabulars and arrays

%% cleveref package for convenient hyperrefercing/citing:
\usepackage[nameinlink,capitalize]{cleveref}
\crefname{equation}{Equation}{Equations}
\Crefname{equation}{Equation}{Equations}
\crefname{figure}{Figure}{Figures}
\Crefname{figure}{Figure}{Figures}

\definecolor[named]{verylightgray}{rgb}{0.93,0.93,0.93}

\input{analysis/fits/modeldefs.tex}
\input{macros.tex}

%% http://tex.stackexchange.com/questions/316510/hyperlink-within-caption-to-a-figure
%% if used in a caption, need short content \caption[...]{ ... \Methodslink ... }
\renewcommand{\Methodslink}{\href[page=8]{ms.pdf}{\emph{Methods}}}

@@

\templatetype{pnassupportinginfo}

\title{Acceleration of plague outbreaks in the second pandemic}

\author{David J.\,D.\ Earn, Junling Ma, Hendrik Poinar, Jonathan Dushoff, Benjamin M.\ Bolker}
\correspondingauthor{E-mail: \texttt{earn@math.mcmaster.ca}}

% Please give the surname of the lead author for the running footer
\leadauthor{Earn}

@@

\maketitle

%% Adds the main heading for the SI text. Comment out this line if you do not have any supporting information text.
\SItext

\section*{Historical human population estimates for London}

See \cref{T:london_pop} and \cref{F:london_pop}.

%
\begin{table}%[h!]
\caption[Population estimates for London]{Population estimates for London previously published by Finlay \cite{Finl81,FinlShea86}\cite[p.\,108]{Kryl11} and Creighton \cite[p.\,660]{Crei65}.  These data are plotted in \cref{F:london_pop}.}
\label{tab:pop}
\begin{center}
%% \input{analysis/tables/supppop.tex} % generated by analysis/tables/supppop.R
%% DE: I'm building the (somewhat different) table myself.
\renewcommand{\arraystretch}{1.0}
\begin{tabular}{c|r|l}
%%\hline
{\bfseries Year} &
{\bfseries Population} &
{\bfseries Source}\\
\hline
  1340 &  45,000 & Finlay \\ 
  1500 &  50,000 & Finlay \\ 
  1550 & 110,000 & Finlay \\ 
  1600 & 180,000 & Finlay \\ 
  1603 & 250,000 & Creighton \\
  1625 & 320,000 & Creighton \\
  1650 & 340,000 & Finlay \\ 
  1665 & 460,000 & Creighton \\
  1700 & 445,000 & Finlay \\ 
\end{tabular}
\end{center}
\label{T:london_pop}
\end{table}
%

%
\begin{figure*}%[h!]
\begin{center}
\includegraphics[width=\textwidth]{analysis/plots/london_pop.pdf}
\end{center}
\caption[Historical population of London]{Historical population of London, England, as estimated by Finlay \cite{Finl81,FinlShea86} (\cref{tab:pop}).  Earlier estimates of Creighton \cite[p.\,660]{Crei65} are shown for the specific plague years 1603, 1625 and 1665.}
\label{F:london_pop}
\end{figure*}
%

%%\FloatBarrier
\section*{Disease-related data sources}

@@

%% requires tabularx package:
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\begin{table*}%[h!]
\caption[]{Disease-related data used in this paper.}
\label{tab:datasources}
\begin{center}
\renewcommand{\arraystretch}{1.5}
\begin{tabular}{c|c|L{3cm}|c|L{3.5cm}}
%%\hline
{\bfseries Epoch} &
{\bfseries Year range} &
{\bfseries Data type} &
{\bfseries Frequency} &
{\bfseries Source} %%&
%%{\bfseries Description}
 \\
\hline
Early & 1340--1380 & Last Wills and Testaments & Daily & Court of Husting \\\hline
Late & 1540--1680 & Last Wills and Testaments & Daily & Prerogative Court of Canterbury (PCC) \\
 & 1540--1680 & Mortality from all causes & Weekly & Parish records\\
 & 1563--1666 & Mortality from plague & Weekly & London Bills of Mortality (LBoM)
\end{tabular}
\end{center}
\end{table*}

\subsection*{Last wills and testaments}

@@

\subsection*{Will dates vs.\ probate dates}

@@

%
\begin{figure*}%[h!]
\begin{center}
  \includegraphics[width=\textwidth]{analysis/plots/probate_dates.pdf}
\end{center}
\caption[Wills written vs.\ probated]{Counts of wills written vs.\ wills probated during each of
  the four 14th century plague epidemics in London, based on wills
  probated in the Court of Husting \cite{Shar1889}. Smooth lines represent loess fits
with span=0.5 (shaded regions are 95\% confidence intervals).}
\label{F:willsVSprobates}
\end{figure*}
%

\subsection*{London Bills of Mortality (LBoM)}

@@

\subsection*{Parish registers}

@@

\subsection*{Major vs minor plague epidemics}

@@

%%\newcommand{\mywidth}{0.8\textwidth}
\newcommand{\mywidth}{\textwidth}
%
\begin{figure*}%[h!]
\begin{center}
\includegraphics[width=\mywidth]{analysis/plots/plotallscaled.crop.pdf}
\end{center}
\caption[Plague deaths per 1000 per week]{Plague deaths per 1000 persons per week, from 1563 to 1666.
Epidemics that exceeded 5 on this scale were classified as \emph{major}.
Analysis of the major epidemics is shown in \cref{tab:params,F:timeseries,F:combdata}.  Corresponding analyses of all of the minor epidemics are presented
in \cref{tab:minorparams,F:seasonalgrowth}.
}
\label{F:all1600s.scaled}
\end{figure*}
%

\begin{figure*}%[h!]
\begin{center}
%%\scalebox{0.6}{
\includegraphics[width=\mywidth]{analysis/plots/plotall.crop.pdf}
%%}
\end{center}
\caption[Plague deaths]{All plague deaths reported in London from 1563 to 1666.}
\label{F:all1600s}
\end{figure*}
%

\begin{figure*}
\begin{center}
\includegraphics[width=0.95\textwidth]{analysis/plots/plotwills14c.pdf}
\end{center}
\caption[]{Weekly numbers of wills written during plague epidemics in London, England, in the 14th century.
Vertical grey lines indicate the ``outbreak window'' specified in \cref{tab:windows}.}
\label{F:plotwills14c}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=0.95\textwidth]{analysis/plots/plot3sourcesmajors.pdf}
\end{center}
\caption[]{Weekly deaths and numbers of wills written during major plague epidemics in London, England, since 1540.  For each epidemic, vertical grey lines indicate the beginning and end of the period during which deaths from plague were listed in the bills of mortality.  The left (right) scale refers to deaths (wills).}
\label{F:plot3sourcesmajors}
\end{figure*}

%%\clearpage
\section*{Generation interval for bubonic plague}\label{sec:gentime}

For pneumonic plague, sufficiently detailed data exist for a number of
modern outbreaks to allow an estimate of the latent and infectious
periods (and hence of the generation interval distribution)
\cite{GaniLeac04}, which we can use to estimate $\R_0$ for a given
value of $r$ \cite{WallLips07}. Much less data is available for bubonic plague; in addition, the more complex host-vector life cycle of bubonic plague complicates the estimation of the generation interval. However, we can say that the elapsed time between the onset of infectiousness of a rat and the time when a rat in the next infection cycle becomes infectious is:
%%
\begin{equation}
  \begin{split}
    &\quad \textrm{rat} \to \textrm{flea infection time} \\
    \mbox{} + &\quad \textrm{max}\bigl( \textrm{flea incubation period},
    \{\textrm{time to rat death} + \textrm{flea searching time}\} \bigr) \\
    \mbox{} + &\quad \textrm{flea} \to \textrm{rat infection time} \\
    \mbox{} + &\quad \textrm{rat incubation period}
  \end{split}
\end{equation}
%%
The second line above takes account of the fact that a flea leaves
its rat host when, and only when, the rat dies \cite{KeelGill00}.  From
\cite{LaPerriere+2009}, we can gather that
\begin{itemize}
\item fleas bite rats $\approx$ 4 times/day; the rat $\to$ flea infection probability is $\approx$ 0.2 and the flea $\to$ rat infection probability is $\approx$ 0.28, suggesting that both the rat $\to$ flea and flea $\to$ rat infection times are $\approx$ 1 day; 
\item the rat infectious period (time until infected rats die) is $\approx$ 4 days (\cite{LaPerriere+2009} cite \cite{Audi03} for this value; \cite{KeelGill00b} give a value of $\approx$ 18 days, which seems unrealistically long --- another source, \cite{Toll+10}, gives values ranging from $\approx$ 4--9 days)
\item the flea incubation period is long (but very variable), ranging from 9--26 days (\cite{LaPerriere+2009}, citing \cite{Poll52})
\item the flea searching time is not explicitly defined by \cite{LaPerriere+2009}, but we guess it is relatively short ($<$ 1--2 days)
\item \cite{LaPerriere+2009} give a value of $\approx$ 1--3 days for the rat incubation period
\end{itemize}
%%
Since the flea incubation period is typically longer than the combination of rat death time and flea searching time (9--26 days vs.\ (4--9 + 1--2) days), we can approximate the generation interval as $\approx$ (1 day) + (9--26 days) + (1 day) + (1--3 days); we use a value of $\gentime=18$ days in the main text.

\section*{Relationship between rat density and $\R_0$ (in rats)}

@@

%
\begin{figure*}%[h!]
\begin{center}
 \includegraphics[width=\textwidth]{analysis/plots/KvsR.pdf}
\end{center}
\caption[Relative change in rat carrying capacity]{Relative change in rat carrying capacity $K_{\rm R}$ as a
  function of relative change in basic reproduction number $\R_0$.
  Solid curves are based on \cref{E:R0ratio}.  The dotted red line is
  based on \cref{E:aKsmall}.  The \Sexpr{round(R0ratio,2)}-fold increase in $\R_0$ that
  requires explanation is indicated in yellow.}
\label{F:KvsR}
\end{figure*}
%

\section*{Effects of rat ecology on growth rates}

@@

As a vector-borne disease, the rate of bubonic plague spread is primarily affected by the flea-rat ratio.
In the traditional Ross-MacDonald model for vector-borne epidemics, $\R_0$ is proportional to the square of vector-host ratio because vectors must independently bite hosts twice (once to become infected and a second time to infect a susceptible host). In the rat-flea model, $\R_0$ is instead proportional to $V/H$ because \emph{all} infected fleas on a host disperse and bite other hosts when their initial host dies.
From Keeling and Gilligan's plague model \cite{KeelGill00b}, the
expected change in $r$ for a change in flea-rat ratio from $K_{\rm
  F0}$ to $K_{\rm F}$ is $(K_{\rm F}/K_{\rm F0})^S$, where $S$ is the
sensitivity ($\approx 1.5$). Thus in order to a see a \foldval
increase in $r$ we would need a change of $4^{(1/S)} \approx \Sexpr{round(4^(1/1.5),1)}$, not in the rat density, but in the number of fleas per rat, which seems unlikely.

As noted in the main text, rat density has an additional, indirect effect on growth rate $r$ and reproduction number $\R_0$: increasing rat density will increase the probability that fleas leaving dying rats will find new, susceptible rat hosts. To crudely estimate the magnitude of this effect, if we again consider Keeling and Gilligan's model \cite{KeelGill00b} then the maximum possible effect of rat density on $\R_0$ by this mechanism would be proportional (if fleas have very low success in finding new rat hosts). That is, in order for $\rzerob$ to increase from $\approx \rzeroearlybest$ in the 14th century to $\approx \rzerolatebest$ in the late epoch, rat density would have to increase by at least $\ratpct\%$. If fleas were already fairly successful at finding hosts in the 14th century, then changes in rat density would be expected to have only a small effect on $\R_0$ in rats.

\section*{Analysis of delays between wills and other sources}

@@

\begin{table*}%[h!]
  \caption[Observed cross-correlation and delay]{Observed cross-correlation and delay between London Bills of Mortality (LBoM) records and other sources. All lags measured in
    weeks.  Rows corresponding to wills data are highlighted in grey.
  }
\begin{center}
  \input{analysis/tables/wills_delay_tab.tex}
\end{center}
\label{tab:delays}
\end{table*}

\section*{Details of numerical optimization}

Robustly fitting phenomenological epidemic curves to small, noisy data
sets proved to be surprisingly challenging\footnote{See ``monsters in
  the basement'',
  \url{https://redpenblackpen.tumblr.com/post/145304820562/monsters-in-the-basement}.}. In
the course of developing the full model we experimented with a range
of optimizers --- specifically the BFGS and Nelder-Mead options for
R's \texttt{optim} function as well as R's \texttt{nlminb} function; we
also tried an approach that iterated back and forth between
Nelder-Mead and BFGS until convergence was achieved or the fit was
sufficiently stable. We chose \texttt{nlminb} because it gave the best
results (highest log likelihood) for point estimates.
In general, we used the standard link functions proposed by Ma \emph{et al.} \cite{Ma+14}, i.e., fitting most of the parameters (growth rate $r$, background mortality/wills rate $b$, final size $K$, Richards shape parameter $s$) on the log scale with a scaled-tangent link ($\eta = \tan(\pi/2 \cdot (x/2-1))$) for the initial condition $x_0$. However, we found that the scaled-tangent link declined too slowly in the tails, leading to numerical instability; we replaced it with a more standard logit link ($\eta = \log(x/(1-x))$).

@@

\FloatBarrier
\clearpage
\section*{Supplementary tables cited in the main text}

\begin{table*}%[h!]
    \cprotect\caption[Summary statistics]{Summary statistics for the model estimating differences across epochs. All parameters are in units of log(growth rate)/year.
      The model includes fixed effects of epoch (early [14th c.]
      vs. late [16th - 17th c.]) and source (wills, parish, London
      Bills of Mortality) and a random effect of outbreak
      year. Variability for each observation is assumed to be
      proportional to the uncertainty in its $\log(r)$ estimate (see
      main text, \emph{Growth rate estimates}). 
      In Wilkinson-Rogers notation, the model formula is: \verb!log.r ~ epoch + source + (1|outbreak.year), disp= ~ 1+offset(log(sdvals^2))!. Parameter estimates are given on the $\log(r)$ scale;
      Wald confidence intervals are given in parentheses.
}  \begin{center}
    \input{analysis/tables/epochsum.tex}
  \end{center}
\label{tab:summarystats}
\end{table*}

\begin{table*}
  \caption[Minor outbreak parameter estimates]{Parameter estimates for minor plague epidemics. Plague epidemics with $\hat r<1/\mathrm{yr}$ or $\hat r>100/\mathrm{yr}$, representing unreliable fits, are excluded; confidence limits that are $<0.1$ or $>200$ are replaced with {\tt NA}. See \Methodslink.}
  \footnotesize
  \rowcolors{2}{white}{verylightgray}
  \begin{center}
    \input{analysis/tables/minorparams.tex}
  \end{center}
  \label{tab:minorparams}
\end{table*}

\begin{table*}
  \caption[Fitting windows]{Outbreak windows and fitting windows used in fitting the major plague epidemics.  See \Methodslink.}
  \rowcolors{2}{white}{verylightgray}
  \input{analysis/tables/suppwindows.tex}
  \label{tab:windows}
\end{table*}

\begin{table*}
  \caption[Fitting windows (minors)]{The start and end dates for the fitting windows used in fitting the minor plague epidemics.  See \Methodslink.}
  \large
  \rowcolors{2}{white}{verylightgray}
  \begin{center}
    \input{analysis/tables/suppwindowsminor.tex}
  \end{center}
  \label{tab:windowsminor}
\end{table*}

\FloatBarrier
\section*{Supplementary figure cited in the main text}

%
\begin{figure*}%[h!]
\begin{center}
\scalebox{0.9}{
  \includegraphics{analysis/plots/seasonalgrowth.pdf}
}
\end{center}
\caption{Growth rate estimates for all the late epoch (1540--1680) London plague epidemics.  The growth rates shown for the major epidemics are the same as those in the right panel of \cref{F:combdata}. 
}
\label{F:seasonalgrowth}
\end{figure*}
%

%%% Add this line AFTER all your figures and tables
%%\FloatBarrier

%%\dataset{dataset_one.txt}{Type or paste legend here.}

%%\dataset{dataset_two.txt}{Type or paste legend here. Adding longer text to show what happens, to decide on alignment and/or indentations for multi-line or paragraph captions.}

\clearpage
\bibliography{plague}
